A pandemic-scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-09-26 latex/main.tex
Pass at the X events section

\begin{table}
\caption{
ARG events associated with Pango X lineages.
Each row for Type I, III and IV events corresponds to the oldest node 
assigned the specified Pango lineage and is the ancestor of all 
nodes assigned that label. Type II events are recombination events
associated with Pango X lineages not meeting Type I criteria.
For Type I and II recombination 
events, the averted shows the number of additional mutations required
under a recombination-free HMM solution. Parent lineages shown in boldface
are nodes exactly matched by sampled sequences. The descendants column
shows the descendant samples broken down by Pango lineage (grouped 
by root lineage, where appropriate).
}
\label{tab:pango_x_lineages}
% scriptsize is quite small here, and we could probably cram it all 
% in with footnotesize if we worked a bit harder on the intercol
% spacing
\begin{scriptsize}
\begin{tabular}{lrrlrll}
\toprule
\multicolumn{7}{l}{Type I event: Recombination coinciding with Pango origin node}\\
&&\multicolumn{2}{c}{Parent lineage} & \multicolumn{2}{c}{Interval}\\
pango & averted & left & right & left & right & descendants \\
\midrule

@@

\midrule
\multicolumn{7}{l}{Type III events: Pango origin nodes derived from 
Type I and II events}\\
pango & name & path len & time & mutations & samples \\
\midrule

@@

\midrule
\multicolumn{7}{l}{Type IV events: non recombinants}\\
pango & muts & parent & samples & nested \\
\midrule

XP  & 1  &&&& \textbf{BA.1.1} & {XP:45} \\
XN  & 1  &&&& \textbf{BA.2} & {XN:120, BA.2:1} \\
XBQ & 1  &&&& CJ.1 & {XBQ:14} \\
XBK & 1  &&&& CJ.1 & {XBK:7} \\
XBE & 3  &&&& BA.5.2 & {XBE:65} \\
XB  & 2  &&&& B.1 & {XB:192} \\
XAZ & 3  &&&& \textbf{BA.5} & {XAZ:133} \\
XAV & 3  &&&& BA.5.1.24 & {XAV:13} \\
XAU & 1  &&&& BA.2 & {XAU:8} \\
XAS & 1  &&&& BA.4 & {XAS:77} \\
XAN & 5  &&&& \textbf{BA.5.1} & {XAN:7} \\
XAJ & 11 &&&& \textbf{BA.2.12} & {XAJ:18} \\

@@

By automatically detecting and incorporating recombination events
into the ARG, sc2ts allows us to model the origins of these 
X lineages as nodes in the graph. 
Table~\ref{tab:pango_x_lineages} summarises
these results for the XX Pango X lineages present in
the ARG. 
For 16 Pango X lineages the oldest node that has been assigned
the focal Pango lineage is a recombination node, and that node 
is the ancestor of all samples assigned that (or a descendant) Pango 
lineage. These are the Type I events, in which the ARG reflects the
expected phylogenetic structure of Pango designation.
Among these events, concordance with the community-consensus 
parent lineages and breakpoint intervals is excellent (see next 
section). 13 of the Type I events are monophyletic for the focal
pango; Figure~\ref{fig:subgraphs}A shows XA as an example.
The remaining 3 Type I events (XBB, XBQ and XM) have descendants not assigned to the 
focal pango, many of which are the generic BA.2 label
[FIXME presumably due to uncertainty? Look up the results CSV? Be precise].

The 5 Type II events in Table~\ref{tab:pango_x_lineages} 
are recombination events closely associated with Pango X lineages, 
but that do not meet the strict origination criteria for Type I.
Type III events are then nodes in the ARG associated with the 
orin

Of these, XAF and XJ are relatively straightforwardo

Furthermore, some of the Pango X lineages above originate from
shared recombination events in the ARG (Table~\ref{tab:pango_x_lineages}).
An example of a set of ``nested'' closely related lineages
consists of XZ, XAC, XAD, XAE, and XAP,
which descend from a single recombination event in the ARG, rather than five independent
ones (Figure~\ref{fig:subgraphs}B), and
which we tentatively group using the unused label Xx.
These Pango X lineages were designated around the same time.
These results highlight that the ARG can help
clarify the relationships between recombinants,
allowing more precise interpretation of evidence for recombination.

Sc2ts identified 12 Pango X lineages in the ARG as non-recombinants
(Table~\ref{tab:pango_x_lineages}), on the basis of the HMM and available data.
For all these Pango X lineages, the number of recombination-informative sites
fell below our threshold of detection (i.e. $k = 4$ in the HMM).
The fact that most of these Pango X lineages (except XAJ and XAN)
differ from a single parent node in the ARG
by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
their origins as non-recombinants.
For example, the closely related XN and XAU can be explained by two mutations
and one mutation in the ARG, respectively (Figure~\ref{fig:subgraphs}C), and we
believe the original recombinant designation to be based on erroneous data
(Supplementary document S1.XXX).
%%% TODO: Say something about XAJ and XAN.
%JK: we should weaken this statement a bit, as we're not certain that they 
% really are recombinants, especially XP.
% YW: we now have a better in-depth breakdown of these recombinants
% to be pasted into doc S1 - e.g. XB is missing the parent lineages, XP is defined on the basis
% of 2 loci, and XAJ on the basis of 3.
However, sc2ts did not detect several Pango X lineages (XB, XP, and XAJ) as recombinants
because the HMM treated characteristic deletions as missing data.

[NOTE: putting this here for now, until we find the right place to slot it
in next draft.]
See Document~\ref{sec:supplement_indepth_pango_x_lineages} for extended analyses,
Figure~\ref{fig:pango_x_copying_patterns} for copying patterns,
and Document S2 for ARG visualations of all Pango X lineages.

@@

\subsection{Analysis of deletions}
\label{sec:supplement_dels}

@@

We also found highly recurrent deletions using the ARG.
The top 20 deletions in the number of occurrences in the ARG
are shown in Table~\ref{tab:rec_dels}.
About half of these deletions are one-base,
occur in non-coding regions, and
are concentrated in the 3' end of the genome.
Four one-base deletions occur
in accessory genes (ORF7a, ORF8, and ORF10).
The two deletions in S1-NTD are known recurrent deletions
associated with antibody escape \cite{McCarthy2021}.

\begin{table}
\centering
\caption{
Major deletion events in the ARG.
These events are inherited by at least 10,000 samples in the ARG.
Count is the number of samples inheriting a deletion.
The Pango lineage assigned to the node
above which an event occurs is shown.
Abbreviations: NSP, non-structural protein;
ORF, open reading frame;
S, Spike;
S1-NTD, N-terminal domain in S1 subunit of Spike.
}
\label{tab:major_dels}
\begin{tabular}{rrcrcrr}
\toprule
Start & Length & Region & Node & Pango lineage & Count & Frequency (\%)\\
\midrule
28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 \\
28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 \\
22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112504 & 44.8200 \\
11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 & 871782 & 35.1220 \\
28362 & 9 & N & 1436802 & B.1.1.529 & 871774 & 35.1216 \\
21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 532072 & 21.4359 \\
11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 531894 & 21.4287 \\
6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 340757 & 13.7283 \\
11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 339867 & 13.6924 \\
21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339681 & 13.6849 \\
21987 & 9 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339651 & 13.6837 \\
22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 312706 & 12.5982 \\
28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 290446 & 11.7014 \\
11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 290408 & 11.6998 \\
21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 288831 & 11.6363 \\
21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 290438 & 11.7010 \\
21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 190425 & 7.6718 \\
686 & 9 & ORF1ab / NSP1 & 2698748 & BA.4 (Omicron) & 23330 & 0.9399 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{
Highly recurrent deletions.
These are the top 20 deletions
in terms of the number of occurrences in the ARG.
Abbreviations:
NSP, non-structural protein;
ORF, open reading frame;
S, Spike;
S1-NTD, N-terminal domain in S1 subunit in Spike.
}
\label{tab:rec_dels}
\begin{tabular}{rrcr}
\toprule
Start & Length & Region & Occurrences \\
\midrule
29781 & 1 & non-coding & 12754 \\
29762 & 1 & non-coding & 11383 \\
29779 & 1 & non-coding & 10951 \\
29769 & 1 & non-coding & 10809 \\
29774 & 1 & non-coding & 10735 \\
29700 & 1 & non-coding & 8562 \\
686 & 9 & ORF1ab / NSP1 & 6436 \\
29614 & 1 & ORF10 & 5039 \\
29555 & 1 & non-coding & 4814 \\
29543 & 1 & non-coding & 4804 \\
29541 & 1 & non-coding & 4802 \\
203 & 1 & non-coding & 3604 \\
222 & 1 & non-coding & 3442 \\
21991 & 3 & S / S1-NTD & 2088 \\
28254 & 1 & ORF8 & 1969 \\
28093 & 1 & ORF8 & 900 \\
21765 & 6 & S / S1-NTD & 482 \\
27532 & 1 & ORF7a & 404 \\
335 & 1 & ORF1ab / NSP1 & 370 \\
337 & 1 & ORF1ab / NSP1 & 346 \\
\bottomrule
\end{tabular}
\end{table}

@@

@@

\subsubsection{XQ/XR/XU/XAA/XAG/XAM}
https://github.com/jeromekelleher/sc2ts-paper/issues/338

>>>>

\begin{table}
\caption{
ARG events associated with Pango X lineages.
Each row for Type I, III and IV events corresponds to the oldest node 
assigned the specified Pango lineage and is the ancestor of all 
nodes assigned that label. Type II events are recombination events
associated with Pango X lineages not meeting Type I criteria.
For Type I and II recombination 
events, the averted shows the number of additional mutations required
under a recombination-free HMM solution. Parent lineages shown in boldface
are nodes exactly matched by sampled sequences. The descendants column
shows the descendant samples broken down by Pango lineage (grouped 
by root lineage, where appropriate).[FIXME Typesetting to be finished,
describe cols for Type III and IV]
}
\label{tab:pango_x_lineages}
% scriptsize is quite small here, and we could probably cram it all 
% in with footnotesize if we worked a bit harder on the intercol
% spacing
\begin{scriptsize}
\begin{tabular}{lrrlrll}
\toprule
\multicolumn{7}{l}{Type I event: Recombination coinciding with Pango origin node}\\
&&\multicolumn{2}{c}{Parent lineage} & \multicolumn{2}{c}{Interval}\\
pango & averted & left & right & left & right & descendants \\
\midrule

@@

\midrule
\multicolumn{7}{l}{Type III events: Pango origin nodes derived from 
Type I and II events}\\
pango & name & path len & time & mutations & & descendants\\
\midrule

@@

\midrule
\multicolumn{7}{l}{Type IV events: non recombinants}\\
pango & muts &&&& parent & descendants \\
\midrule

XAS & 1  &&&& BA.4 & {XAS:77} \\
XAU & 1  &&&& BA.2 & {XAU:8} \\
XBK & 1  &&&& CJ.1 & {XBK:7} \\
XBQ & 1  &&&& CJ.1 & {XBQ:14} \\
XN  & 1  &&&& \textbf{BA.2} & {XN:120, BA.2:1} \\
XP  & 1  &&&& \textbf{BA.1.1} & {XP:45} \\
XB  & 2  &&&& B.1 & {XB:192} \\
XAV & 3  &&&& BA.5.1.24 & {XAV:13} \\
XAZ & 3  &&&& \textbf{BA.5} & {XAZ:133} \\
XBE & 3  &&&& BA.5.2 & {XBE:65} \\
XAN & 5  &&&& \textbf{BA.5.1} & {XAN:7} \\
XAJ & 11 &&&& \textbf{BA.2.12} & {XAJ:18} \\

@@

By automatically detecting and incorporating recombination events
into the ARG, sc2ts provides us with a systematic means of 
discovering and assessing evidence around these events.
Table~\ref{tab:pango_x_lineages} summarises
these results for the 44 Pango X lineages present in
the ARG (XAC and XAD have been omitted, and XM 
partially represented, corresponding to 23 samples in total; see STAR Methods).
For 16 Pango X lineages the oldest node that has been assigned
the focal Pango lineage is a recombination node, and that node 
is the ancestor of all samples assigned that (or a descendant) Pango 
lineage. These are the Type I events, in which the ARG reflects the
expected phylogenetic structure of Pango designation.
Among these events, concordance with the community-consensus 
parent lineages and breakpoint intervals is excellent (see next 
section). 13 of the Type I events are monophyletic for the focal
pango (i.e., the descendants are entirely from that lineage); 
Figure~\ref{fig:subgraphs}A shows XA as an example (see also 
Document~\ref{sec:jackson_recombs} for further analysis 
of the Jackson et al.\ recombinants).

The 5 Type II events in Table~\ref{tab:pango_x_lineages} 
are recombination events closely associated with Pango X lineages, 
but that do not meet the strict origination criteria for Type I.
Type III events are Pango origin nodes in the ARG that are 
derived from Type I or Type II recombination events. There is significant
nesting among the Pango X lineage origin nodes within the ARG,
with two major clusters of particular note. Although XQ is 
directly associated with a recombination event in the ARG, 
XAA, XAG, XAM, XR and XU are all closely derived, suggesting 
that they are more parsimoniously explained as sublineages of XQ
than de-novo recombinant lineages.
The second cluster is based around a putative undocumented
recombinant that is the ancestor of lineages 
XZ, XAC, XAD, XAE, and XAP, which we refer to as ``Xx''
(Figure~\ref{fig:subgraphs}B).

The final class of event (Type IV) in Table~\ref{tab:pango_x_lineages} 
correspond to nodes that are not associated with a recombination
event. These vary in their plausibility of being non-recombinant.
For example, XN (Figure~\ref{fig:subgraphs}C) is only one mutation
distant from a BA.2 sample and has little additional evidence to 
suggest recombination [ref sup section]. XB, on the other hand,
has been thoroughly characterised\cite{Gutierrez2022} and 
may have misclassified by sc2ts due to limitations in handling
deletions or matching against long branches (see Discussion).
See Document~\ref{sec:supplement_indepth_pango_x_lineages} for extended 
analyses on all of the Pango X lineages above,
Figure~\ref{fig:pango_x_copying_patterns} for copying patterns 
associated with Table~\ref{tab:pango_x_lineages},
and Document S2 for ARG visualisations.

The nested samples not assigned to the focal pango lineage are
either from different pango X lineages (see next paragraph), 
or given a generic BA.2 assignment. Of these 59 generic assignents
6 had conflicts in their UShER placements and could have been
assigned an X lineage (XJ: 3, XU: 2, XN: 1), and 
25 were cases in which Scorpio~\cite{Colquhoun2023scorpio}
failed to find a specific lineage (accounting for 7\%
of the 357 ``Omicron (Unassigned)`` Scorpio calls across
2,747,985 alignments).

@@

\subsection{Analysis of deletion events}
\label{sec:supplement_dels}

@@

We also found highly recurrent deletions using the ARG.
The top 20 deletions in the number of occurrences in the ARG
are shown in Table~\ref{tab:rec_dels}.
About half of these deletions are one-base and
occur in the non-coding regions in the 3' end of the genome
between the accessory genes.
Four one-base deletions occur
in the accessory genes ORF7a, ORF8, and ORF10.
The two deletions in S1-NTD are known recurrent deletions
associated with antibody escape \cite{McCarthy2021}.

\begin{table}
\centering
\caption{
Major deletion events in the ARG.
These events are inherited by
at least 10,000 samples in the ARG.
Count is the number of samples inheriting a deletion.
The Pango lineage assigned to the node
above which an event occurs is shown.
Abbreviations: NSP, non-structural protein;
ORF, open reading frame;
S, Spike;
S1-NTD, N-terminal domain in Spike S1 subunit.
}
\label{tab:major_dels}
\begin{tabular}{rrcrcrr}
\toprule
Start & Length & Region & Node & Pango lineage & Count & Frequency (\%)\\
\midrule
28271 & 1 & non-coding & 1436808 & B.1 & 1118363 & 45.0561 \\
28248 & 6 & ORF8 & 220186 & B.1.617.2 (Delta) & 1114669 & 44.9073 \\
22029 & 6 & S / S1-NTD & 200039 & B.1.617.2 (Delta) & 1112504 & 44.8200 \\
11288 & 4 & ORF1ab / NSP6 & 1436802 & B.1.1.529 (Omicron) & 871782 & 35.1220 \\
28362 & 9 & N & 1436802 & B.1.1.529 (Omicron) & 871774 & 35.1216 \\
21633 & 9 & S / S1-NTD & 822854 & BA.2 (Omicron) & 532072 & 21.4359 \\
11292 & 5 & ORF1ab / NSP6 & 822854 & BA.2 (Omicron) & 531894 & 21.4287 \\
6513 & 3 & ORF1ab / NSP3 & 851246 & BA.1 (Omicron) & 340757 & 13.7283 \\
11283 & 5 & ORF1ab / NSP6 & 851246 & BA.1 (Omicron) & 339867 & 13.6924 \\
21765 & 6 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339681 & 13.6849 \\
21987 & 9 & S / S1-NTD & 851246 & BA.1 (Omicron) & 339651 & 13.6837 \\
22194 & 3 & S / S1-NTD & 851246 & BA.1 (Omicron) & 312706 & 12.5982 \\
28271 & 1 & non-coding & 86456 & B.1.1.7 (Alpha) & 290446 & 11.7014 \\
11288 & 9 & ORF1ab / NSP6 & 86456 & B.1.1.7 (Alpha) & 290408 & 11.6998 \\
21991 & 3 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 288831 & 11.6363 \\
21765 & 6 & S / S1-NTD & 86456 & B.1.1.7 (Alpha) & 290438 & 11.7010 \\
21765 & 6 & S / S1-NTD & 1265302 & BA.4 (Omicron) & 190425 & 7.6718 \\
686 & 9 & ORF1ab / NSP1 & 2698748 & BA.4 (Omicron) & 23330 & 0.9399 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{
Highly recurrent deletions.
These are the top 20 deletions
in terms of the number of occurrences in the ARG.
Abbreviations:
NSP, non-structural protein;
ORF, open reading frame;
S, Spike;
S1-NTD, N-terminal domain in Spike S1 subunit.
}
\label{tab:rec_dels}
\begin{tabular}{rrcr}
\toprule
Start & Length & Region & Occurrences \\
\midrule
29781 & 1 & non-coding & 12754 \\
29762 & 1 & non-coding & 11383 \\
29779 & 1 & non-coding & 10951 \\
29769 & 1 & non-coding & 10809 \\
29774 & 1 & non-coding & 10735 \\
29700 & 1 & non-coding & 8562 \\
686 & 9 & ORF1ab / NSP1 & 6436 \\
29614 & 1 & ORF10 & 5039 \\
29555 & 1 & non-coding & 4814 \\
29543 & 1 & non-coding & 4804 \\
29541 & 1 & non-coding & 4802 \\
203 & 1 & non-coding & 3604 \\
222 & 1 & non-coding & 3442 \\
21991 & 3 & S / S1-NTD & 2088 \\
28254 & 1 & ORF8 & 1969 \\
28093 & 1 & ORF8 & 900 \\
21765 & 6 & S / S1-NTD & 482 \\
27532 & 1 & ORF7a & 404 \\
335 & 1 & ORF1ab / NSP1 & 370 \\
337 & 1 & ORF1ab / NSP1 & 346 \\
\bottomrule
\end{tabular}
\end{table}

@@

\subsection{Major saltation lineages}

% https://github.com/jeromekelleher/sc2ts-paper/issues/772

@@

\subsubsection{XQ/XR/XU/XAA/XAG/XAM}
\label{sec:supplement_indepth_pango_x_lineages_xq}
https://github.com/jeromekelleher/sc2ts-paper/issues/338
